---
title: "Paz_Thal_MG_Clustering_2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
#install.packages("BiocManager")
#BiocManager::install("Seurat")
#BiocManager::install("ggplot2")
#install.packages("dplr")
#install.packages("ape")
#install.packages("cowplot")
#install.packages("Matrix")
```

Load packages into your workspace. 

```{r}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ape)
library(cowplot)
library(Matrix)

#find out where you are
getwd()
```

*Edit the following code every time* 
and make sure the folders "QC" and "Plots" and "Heatmaps" and "Trees" and "Spreadsheets" and "Data/Seurat" are present in the "dir" folder
```{r}
#Specify your preferred directory for all input + output
dir= "/Users/whippoorwill/Desktop/Sequencing/Paz_Thalamus_2022/"

#Specify exactly where your seurat files live
datafolder = "Data/Seurat"

#set up folders
QCfolder = "QC"
Plotfolder = "Plots"

#This name needs to match your project name within the seurat object
project<-"Microglia_Paz_MG_Thal"


#the data file you want to open
filename = paste0(project,"_filtered.RData")

#Specify your organism; please capitalize the first letter (i.e. "Mouse", "Human","Zebrafish","Rat")
organism = "Mouse"

#metrics you want to look at for QC
m = c("nCount_RNA","nFeature_RNA","percent.mito","percent.ribo")

#You can add in housekeeping genes if you want them to be scaled always; otherwise set to "NULL"
add.genes = NULL

#Choose what to regress out - could be age, sex, or any metadata column
regress = c("percent.mito","nFeature_RNA")

#Decide whether to use SCtransform (can take a long time with big datasets; generally recommended)
sct = TRUE

#How many genes do you want scaled to use for UMAP/clustering? 
ngenes = 2000

#Which principal components do you want to calculate on? This is a default setting, change if one of the pc's is for something you think is a technical error (i.e. HSP, RP, etc)
pcs = c(1:30)

#clustering resolution; the last number will be saved as "seurat_clusters" in metadata
res = c(1.5,1.0,0.5)

#Important genes to determine your cells of interest
igenes = c("Slc1a3","Pecam1","Rbfox3","Pdgfra","Mbp","Cspg4","P2ry12","Hexb","Ptprc")

#metadata dimensions you want to cluster on
dims = c("seurat_clusters","sample_description","treatment","condition","fulldescription")

#make a unique name; maybe this is the celltype you've already subset, or the age you're looking at, etc. 
iterationname = "mg1"

#Establish cutoffs for heatmaps
pval = 1e-8 #max p-value for significance
lfc = 0.2 #minimum log fold change
minpct = 0 #if you want to subset by percent cells in that cluster expressing the gene
maxpct = 1
single = F #should each gene be a marker of a single cluster only
hgenes = 3 #how many genes should be in the heatmap per cluster
ncells = 100 #max # of cells per heatmap column
column = "seurat_clusters" #division you care about

#Variance partition: Remove genes only associated with a technical variable or sex of the mouse
variance = F
```


Load in your filtered dataset
```{r}
load(file.path(dir,datafolder,filename))
```
```{r}
table(sobject$fulldescription)
```

Function to print multiple graphs: 
```{r}
PrintSeuratGraph = function(namecard = "a",seurat_object = sobject,graphtype = "feature",feature = NULL,group = NULL,split=NULL,cellnames=NULL){
  if (!is.null(cellnames)){
    Idents(seurat_object) = cellnames[1]
    cells = colnames(seurat_object)[Idents(seurat_object) %in% cellnames[2:length(cellnames)]]} 
  else {cells = cellnames}
  if (graphtype == "feature"){
    graph = FeaturePlot(seurat_object,features = feature,split.by = split, cells = cells,cols = c("lightyellow","darkred"))
  }
  if (graphtype == "violin"){
    graph = VlnPlot(seurat_object,features = feature, pt.size = 0.1, idents = cellnames[2:length(cellnames)],group.by = group, split.by = split) + stat_summary(fun = median, geom='point', size = 10, colour = "black", shape = 95) + NoLegend()
  }
  if (graphtype == "dim"){
    graph = DimPlot(seurat_object,cells = cells, group.by = group, split.by = split)
    
  }
  name = paste0(feature,"_",graphtype,namecard,".eps")
  graph
  setEPS()
  postscript(file.path(dir,Plotfolder,name))
  print(graph)
  dev.off()
}
```



Find variable features, normalize, scale, run PCA, clustering, umap
The following is the standard method of normalization and scaling. Interchangeable with the next chunk. Run both, you should have already specified which to use with "sct = T or F". Will take 5-20 minutes to run. 


Alternative: SCTransform (great for smaller datasets)
```{r, results='hide',eval = FALSE}
if (sct){
  sobject <- SCTransform(sobject, vars.to.regress = regress, verbose = FALSE,variable.features.n = ngenes,conserve.memory = T)
}
```

******

Show most variable genes
```{r}
labels <- c(head(VariableFeatures(sobject),10),add.genes)
plot1 = VariableFeaturePlot(sobject)
LabelPoints(plot=plot1, points = labels, repel = F, xnudge = 0.1, ynudge = 0.5)
```

Remove genes that correlate strongly with technical variables from downstream analyses

```{r}
sexgenes = c("Xist","Tsix","Ddx3y","Eif2s3y","Uty","Fkbp5", "Ddit4")
var.genes = VariableFeatures(sobject)
var.genes = var.genes[!var.genes %in% sexgenes]
VariableFeatures(sobject) = var.genes
```
*****
Run PCA analysis and show elbow plot
```{r}
sobject <- RunPCA(sobject,features = VariableFeatures(sobject),npcs = 50, verbose = FALSE)
ElbowPlot(sobject,ndims = 50, reduction = "pca")
print(sobject[["pca"]], dims = 1:20, nfeatures = 5)
```

Once you are satisfied with pc's, run clustering: 
```{r}
sobject<-RunUMAP(sobject,reduction = "pca",dims = pcs, verbose = F)
sobject<-FindNeighbors(sobject,dims=pcs,verbose=F)
sobject<-FindClusters(sobject,verbose=F,resolution = 0.5)
```

Plot important objects; check parameters before moving forward, evaluate QC, clustering
```{r}
for (dim in dims){
  print(DimPlot(sobject,group.by = dim, label = T))
}

FeaturePlot(sobject,igenes)
FeaturePlot(sobject,m)

VlnPlot(sobject,igenes,sort = "increasing",pt.size = 0.01)

#Build a clustering tree
Idents(sobject) = "seurat_clusters"
sobject= BuildClusterTree(sobject,dims = pcs)
tree = sobject@tools$BuildClusterTree
plot.phylo(tree, use.edge.length = T, direction = "rightwards")
```
****
Note: delete cluster 5, hashed because I only ran this the first time
```{r}
#sobject = subset(sobject,subset = seurat_clusters != "5")
#save(sobject,file = file.path(dir,datafolder,filename))
```

If you are happy with the QC, you can move on to either subdividing further to isolate interesting cells (below) or running differential expression comparisons on the existing clustering. 

*Note that the chunk below requires you to manually decide which clusters to keep or exclude. The violin plots will help decide based on marker genes. You can also choose to annotate (as many types as you want) and skip subsetting.*

*****
Save the tree
```{r}
Idents(sobject) = column
sobject= BuildClusterTree(sobject,dims = pcs)
tree = sobject@tools$BuildClusterTree
setEPS()
postscript(file.path(dir,"Trees",paste0(project,"_tree_",iterationname,".eps")))
plot.phylo(tree, use.edge.length = T, direction = "rightwards")
dev.off()

plot.phylo(tree, use.edge.length = T, direction = "rightwards")

VlnPlot(sobject,"percent.mito",group.by = "seurat_clusters")+NoLegend()
```

Save the clustered dataset. Overwrite the existing, subset dataset. 
```{r,eval = FALSE}
save(sobject,file = file.path(dir,datafolder,filename))
```
*************************************


Block to print multiple graphs: 
```{r}
name = paste0(project,iterationname)
genes = igenes
features = m
groups = c(dims,"celltype","celltypecluster")
genes = genes[genes %in% rownames(GetAssayData(sobject,slot = "data"))]

for(feature in genes){
  PrintSeuratGraph(namecard = name,graphtype = "feature",feature = feature)
}

for(feature in features){
  PrintSeuratGraph(namecard = name,graphtype = "feature",feature = feature)
}

#split feature plots by individual
for(feature in c(features)){
  PrintSeuratGraph(namecard = paste0(name,"_split"),graphtype = "feature",feature = feature,split = "sample_description")
}

#dim plots for clustering
for(group in groups){
  PrintSeuratGraph(namecard = name,graphtype = "dim",group = group, feature = group)
}

#violin plots
for(feature in c(genes,features)){
  PrintSeuratGraph(namecard = name,graphtype = "violin",feature = feature,group = "seurat_clusters")
}
```

*Heatmap*

First calculate DE genes for every cluster
```{r,results = "hide",eval = FALSE}
Idents(sobject) = column
markers_all <- FindAllMarkers(
    object = sobject,
    features = VariableFeatures(sobject),
    assay = "RNA",
    slot = "counts",
    test.use = "DESeq2",
    only.pos = FALSE, 
    min.pct = 0.05, 
    logfc.threshold = 0.2
)
```


```{r,results = "hide",eval = FALSE}
write.csv(markers_all,file = file.path(dir,"Spreadsheets",paste0(iterationname,"_all_markers.csv")))
```

Make the heatmap
```{r}
#read in a de gene file
markers = read.csv(file.path(dir,"Spreadsheets",paste0(iterationname,"_all_markers.csv")),stringsAsFactors = F)

#Select only the genes that pass thresholds
markers = markers[markers$p_val_adj < pval,]

#pick only positives, or restrict by min/max pct expression using pct1/2
markers = markers[markers$avg_log2FC > lfc,] 
markers = markers[markers$pct.1 > minpct & markers$pct.2 < maxpct,]

#If you want, select markers that define a single cluster
if (single){markers <- markers[markers$gene %in% names(table(markers$gene))[table(markers$gene) == 1],] }

table(markers$cluster)

topgenes <- markers %>% group_by(cluster) %>% top_n(hgenes, avg_log2FC)
topgenes = topgenes[order(topgenes$cluster),]

#Subset each cluster to ncells
cellnames = sobject@meta.data[,column]
names(cellnames) = colnames(sobject)
clusters = levels(as.factor(cellnames))
newcellnames = NULL
for (cluster in clusters){
  n = length(cellnames[cellnames == cluster])
  if (n > ncells){n = ncells}
  newcluster = sample(cellnames[cellnames == cluster],n, replace = F)
  newcellnames = c(newcellnames,newcluster)
}

#check
table(newcellnames)

#Make heatmap
setEPS()
postscript(file.path(dir,"Heatmaps", paste0(iterationname,"_",column,"_",pval,"_ncells",ncells,"heatmap.eps")))
DoHeatmap(
    object = sobject, 
    features = c(topgenes$gene),
    cells = names(newcellnames),
    group.by = column,
    size = 5,
    label = T,
    draw.lines = T
)

dev.off()
```

Print heatmap to console

```{r}
DoHeatmap(
    object = sobject, 
    features = c(topgenes$gene),
    cells = names(newcellnames),
    group.by = column,
    size = 5,
    label = T,
    draw.lines = T
)
```

```{r}
genes = unique(topgenes$gene)
genes
DotPlot(sobject,features = genes,group.by = "seurat_clusters",cluster.idents = T)+RotatedAxis()
```
```{r}
barres = read.csv(file.path(dir,"Paz_MG_TBI/Spreadsheets/barreslab_rnaseq.csv"))
celltype = "Neuron"
column = 4
barres = cbind(barres, "NeuralEnrichment" = rep(0,nrow(barres)))
for (n in 1:nrow(barres)){
  row = barres[n,]
  row[c(3:9)] = as.numeric(row[c(3:9)])
  max = max(barres[n,c(3,5:9)])
  neuralenrichment = barres[n,4]/max
  barres[n,"NeuralEnrichment"] = neuralenrichment
}
neural = barres[barres$NeuralEnrichment>2,]
neuralgenes = neural$Gene.symbol
neuralgenes = neuralgenes[neuralgenes %in% rownames(sobject)]
neurogenes=list()
neurogenes[[1]] = neuralgenes
sobject = AddModuleScore(sobject,features = neurogenes,name = "neurogenes")

damgenes = read.csv("/Users/whippoorwill/Desktop/Sequencing/IL33_scRNAseq_thalamus/mmc3.csv")
head(damgenes)
damgenes = damgenes[damgenes$X.log10.DAM..p.value..Mann.Whitney.>5,]
dg = damgenes[damgenes$Fold.change..DAM.to.homeostatic.microglia.>1.5,]
dg = dg[!is.na(dg$Gene.name),]
dg = dg$Gene.name
dg = dg[dg %in% rownames(sobject)]
dglist = list()
dglist[[1]]=dg
sobject=AddModuleScore(sobject,features = dglist,name = "DAM")

ifitgenes = read.csv("/Users/whippoorwill/Desktop/Sequencing/LD_AVM02/Spreadsheets/allmarkers_vargenesMG_0501 copy 2.csv",stringsAsFactors = F)
head(ifitgenes)
ifitgenes = ifitgenes[ifitgenes$cluster == '8',]
ifitgenes = ifitgenes[ifitgenes$p_val_adj<1e-5,]
ig = ifitgenes$gene[ifitgenes$avg_logFC>0.23]
head(ig)
length(ig)
ig = ig[ig %in% rownames(sobject)]
iglist = list()
iglist[[1]]=ig

sobject=AddModuleScore(sobject,features = iglist,name = "IFN")

mhc = c("H2-Ab1","H2-Aa","H2-Eb1","H2-Eb2") #H2-Ea
mhclist = list(mhc)
sobject = AddModuleScore(sobject,features = mhclist,name = "MhcII")

VlnPlot(sobject,"IFN1",group.by = "seurat_clusters",pt.size = 0)+NoLegend()
VlnPlot(sobject,"MhcII1",group.by = "seurat_clusters",pt.size = 0)+NoLegend()
VlnPlot(sobject,"DAM1",group.by = "seurat_clusters",pt.size = 0)+NoLegend()
VlnPlot(sobject,"neurogenes1",group.by = "seurat_clusters",pt.size = 0)+NoLegend()
```





```{r}
t = table(sobject$seurat_clusters,sobject$sample_description)
b = apply(t,MARGIN = 2,function(x)x/sum(x)*100)
b = apply(b,MARGIN = 1,function(x)x/sum(x)*100)
b
barplot(b,beside = T,legend = rownames(b),col=c("blue","black","red"),ylim = c(0,150),xlim = c(0,50)) + abline(h=33)

setEPS()
postscript(file.path(dir,"Plots",paste0(project,"_barplot_normalized2x.eps")))
barplot(b,beside = T,legend = rownames(b),col=c("blue","black","red"),ylim = c(0,150),xlim = c(0,50)) + abline(h=33)
dev.off()
```


```{r,eval=F,results = 'hide'}
save(sobject,file = file.path(dir,datafolder,filename))
```

```{r}
# install from CRAN
#install.packages("gprofiler2")
# load the package
library(gprofiler2)

#read in a de gene file
markers = read.csv(file.path(dir,"Spreadsheets",paste0(iterationname,"_all_markers.csv")),stringsAsFactors = F)

#Select only the genes that pass thresholds
markers = markers[markers$p_val_adj<pval,]

#pick only positives, or restrict by min/max pct expression using pct1/2
markers = markers[markers$avg_log2FC > lfc,] 
markers = markers[markers$pct.1 > minpct & markers$pct.2 < maxpct,]

for (cluster in names(table(markers$cluster))){
  deg = markers[markers$cluster == cluster,]
  gp_up = gost(deg$gene, organism = "mmusculus",evcodes = T,sources = c("GO","KEGG"), custom_bg = rownames(sobject))
  
  if (!is.null(gp_up)){
    results = as.data.frame(gp_up$result)
    results = apply(results,2,as.character)
    write.csv(results,file = file.path(dir,"Spreadsheets",paste0("gprofiler_upregulated_cluster_",cluster,".csv")))
    print(cluster)
    print(gostplot(gp_up, interactive = TRUE))

  }
  
}
```
****

```{r}
FeaturePlot(sobject,c("Hexb","Cd68","P2ry12","Ifitm3","Spp1","H2-Ab1"),ncol = 3)
VlnPlot(sobject,c("Hexb","Cd68","P2ry12","Ifitm3","Spp1","H2-Ab1"),group.by = "seurat_clusters")

allcells = sobject$sample_description
cellnames = c()
for (s in names(table(sobject$sample_description))){
  newcells = sample(allcells[allcells == s],2500)
  newcells = newcells
  cellnames = c(cellnames,newcells)
}
table(cellnames)
newobject = subset(sobject,cells = names(cellnames))
DimPlot(newobject,group.by = "seurat_clusters",split.by = "sample_description")

VlnPlot(sobject,"Irf1",group.by = "sample_description")
VlnPlot(sobject,"Stat1",group.by = "sample_description")
VlnPlot(sobject,c("H2-K1","H2-Ke6","H2-Oa","H2-DMa","H2-DMb2","H2-DMb1","H2-Ob","H2-Aa","H2-Eb1"),group.by = "sample_description")
VlnPlot(sobject,"Rela",group.by = "sample_description")
```

```{r,eval = F, results = 'hide'}
Idents(sobject) = "sample_description"
markers = FindMarkers(sobject,ident.1 = "TBI-aCD4",ident.2 = "TBI-IgG",group.by = "sample_description",assay = "RNA",slot = "counts",test.use = "MAST",min.pct = 0.05,logfc.threshold = 0.0)
markers$gene = rownames(markers)
write.csv(markers,file = file.path(dir,"Spreadsheets","microglia_CD4vsIgG_TBI_markers.csv"))
```


```{r}
markers = read.csv(file = file.path(dir,"Spreadsheets","microglia_CD4vsIgG_TBI_markers.csv"))
markers = markers[markers$p_val_adj<pval,]
markers = markers[order(markers$avg_log2FC,decreasing = T),]
topgenes = markers$X[1:10]
topgenes = c(topgenes,markers$X[(nrow(markers)-10):nrow(markers)])
```


```{r,eval = F, results = 'hide'}
markers = FindMarkers(sobject,ident.1 = "TBI-IgG",ident.2 = "Control-IgG",group.by = "sample_description",assay = "RNA",slot = "counts",test.use = "MAST",min.pct = 0.05,logfc.threshold = 0.0)
markers$gene = rownames(markers)
write.csv(markers,file = file.path(dir,"Spreadsheets","microglia_TBIvsControl_markers.csv"))
```


```{r}
markers = read.csv(file = file.path(dir,"Spreadsheets","microglia_TBIvsControl_markers.csv"))
markers = markers[markers$p_val_adj<pval,]
markers = markers[order(markers$avg_log2FC,decreasing = T),]
topgenes = c(markers$X[1:10],topgenes)
topgenes = c(topgenes,markers$X[(nrow(markers)-10):nrow(markers)])

```


```{r}
DotPlot(sobject,features = unique(topgenes),group.by = "sample_description",cluster.idents = T)+RotatedAxis()

```
```{r}

allcells = sobject$sample_description
cellnames = c()
for (s in names(table(sobject$sample_description))){
  newcells = sample(allcells[allcells == s],100)
  newcells = newcells
  cellnames = c(cellnames,newcells)
}

DoHeatmap(
    object = sobject, 
    features = topgenes,
    cells = names(cellnames),
    group.by = "sample_description",
    size = 5,
    label = T,
    draw.lines = T
)
```

